Pigeons.jl: Distributed sampling from intractable distributions

### 2024-05-20 paper/PT.tex
Add weak scaling plots

\subsection{Strong and weak scaling of PT}

We study the strong and weak scaling properties of Pigeons' implementation of PT
as given by Amdahl's law and Gustafson's law, respectively.
For the strong scaling results, we consider a fixed problem size and compare the 
compute time as we vary the number of processes.  
For weak scaling, we consider a varying problem size proportional 
to the number of processes used. 
In both cases we use the number of chains as the definition of the problem size; 
this is in line with previous heuristics that establish that the generaly problem 
difficulty is proportional to the optimal number of chains in PT \cite{syed2021nrpt}.

@@

\medskip
The results for strong and weak scaling are presented in \cref{fig:PT_scaling}. 
For strong scaling, we see that ...
For weak scaling, We see that ...
\NS{Comment on plots}.

\begin{figure*}[t]
    \centering
    \begin{minipage}{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/trace_density_plots_pt.pdf}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/trace_density_plots_single_chain.pdf}
    \end{minipage}
    \caption{
        \NS{Add appropriate plots.}
        Strong and weak scaling results for PT.  
        \textbf{Left:} Strong scaling. 
        \textbf{Right:} Weak scaling. 
    }
    \label{fig:PT_scaling}
\end{figure*}

>>>>

\subsection{Weak scaling of PT}

We study the weak scaling properties of Pigeons' implementation of PT
as given by Gustafson's law.
In this setup, we consider a varying problem size proportional to the number of processes used. 
The number of chains is our definition of the problem size; 
this is in line with previous heuristics that establish that the problem 
difficulty is generally proportional to the optimal number of chains in PT \cite{syed2021nrpt}.
In particular, we consider a multivariate normal target distribution 
and vary the dimension $d \in \cbra{2^1, 2^2, \ldots, 2^{12}}$ for the distributed setup and 
$d \in \cbra{2^1, 2^2, \ldots, 2^{10}}$ for the parallel (shared memory) setup. 
The number of chains is set to $\ceil{\sqrt{d}}$ according to PT tuning suggestions \cite{syed2021nrpt} 
and for the distributed and parallel settings we set the number of processes or threads equal 
to the number of chains.
Within each chain we use the autoMALA sampler \cite{biron2024automala}, which is a 
gradient-based sampler with automatic step size selection in the Metropolis-adjusted Langevin algorithm.
Ten different seeds are used in order to obtain approximate 95\% confidence intervals of 
the speedup.

@@

\medskip
The weak scaling results for both the DMP and SMP settings are presented in \cref{fig:PT_scaling}. 
We see that we obtain an approximately linear speedup as we increase the number of processes.

\begin{figure*}[t]
    \centering
    \begin{minipage}{0.4\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/speedup_distributed.pdf}
    \end{minipage}
    \begin{minipage}{0.4\textwidth}
      \centering
      \includegraphics[width=\textwidth]{../img/speedup_parallel.pdf}
    \end{minipage}
    \caption{
        Weak scaling results for PT in both the distributed (MPI) and parallel settings 
        with approximate 95\% confidence intervals for the speedup.  
        The blue dotted line represents a hypothetical speedup equal to the 
        number of processes/threads requested. 
        The red dashed line indicates a speedup of one for reference.
        \textbf{Left:} Distributed setup. 
        \textbf{Right:} Parallel (shared) setup. 
    }
    \label{fig:PT_scaling}
\end{figure*}

### 2024-05-20 paper/examples.tex
Add weak scaling plots

\section{Examples}
In this section we present a set of minimal examples 
that demonstrate how to use Pigeons.jl for sampling. 
For further reading, we also direct readers to our growing list of examples, which can be found at 
\url{https://github.com/Julia-Tempering/Pigeons.jl/tree/main/examples}.
\NS{Add InferenceHub description and link.}

@@

From here, it is straightforward to sample from the density given by 
\cref{eq:coinflip_density} up to a normalizing constant.
We use non-reversible parallel tempering \cite{syed2021nrpt} (PT), Pigeons.jl's 
state-of-the-art sampling algorithm, to sample from the target distribution.
PT comes with several tuning parameters and \cite{syed2021nrpt} describe how 
to select these parameters effectively, which Pigeons.jl implements under the hood.
We also specify that we would like to store 
the obtained samples in memory to be able to produce trace-plots, as well 
as some basic online summary statistics of the target distribution and useful 
diagnostic output by specifying 
\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.allocation\_extrema]}.
It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
will be selected for the user.
The code below runs Pigeons.jl on one machine with one thread.
We use the default values for most settings, however 
we explain later how one can obtain improved performance by setting 
arguments more carefully (see \cref{sec:additional_options}).
\begin{lstlisting}[language = Julia]
using Pigeons
pt = pigeons(
    target = TuringLogPotential(model), 
    record = [
        traces, online, round_trip, 
        Pigeons.timing_extrema, 
        Pigeons.allocation_extrema])
\end{lstlisting}
Note that to convert the Turing model into an appropriate Pigeons.jl target for sampling, 
we pass the model as an argument to \texttt{TuringLogPotential()}.
Once we have stored the PT output in the variable \texttt{pt} we can 
access the results, as described in the following section. 
The standard output after running the above code chunk is displayed in \cref{fig:standard_out}
and explained in the next section. 
For purposes of comparison, we also run a traditional 
(single-chain Markov chain Monte Carlo) method.

>>>>

\section{Examples}
In this section we present a set of minimal examples 
that demonstrate how to use Pigeons.jl for sampling. 
We also direct readers to our growing list of examples at InferHub 
(\url{https://julia-tempering.github.io/InferHub/}), which hosts a growing collection of posterior 
distributions with an emphasis on difficult (non-log-concave) problems.

@@

From here, it is straightforward to sample from the density given by 
\cref{eq:coinflip_density} up to a normalizing constant.
We use non-reversible parallel tempering \NS{add newest ergodicity cite?} 
\cite{syed2021nrpt,syed2021paths,surjanovic2022vpt} (PT), Pigeons.jl's 
state-of-the-art sampling algorithm, to sample from the target distribution.
PT comes with several tuning parameters and \cite{syed2021nrpt} describe how 
to select these parameters effectively, which Pigeons.jl implements under the hood.
We also specify that we would like to store 
the obtained samples in memory to be able to produce trace-plots, as well 
as some basic online summary statistics of the target distribution and useful 
diagnostic output by specifying 
\texttt{record = [traces, online, round\_trip, Pigeons.timing\_extrema, Pigeons.allocation\_extrema]}.
It is also possible to leave the \texttt{record} argument empty and reasonable defaults 
will be selected for the user.
The code below runs Pigeons.jl on one machine with one thread.
We use the default values for most settings, however 
we explain later how one can obtain improved performance by setting 
arguments more carefully (see \cref{sec:additional_options}).
\begin{lstlisting}[language = Julia]
using Pigeons
pt = pigeons(
    target = TuringLogPotential(model), 
    record = [
        traces, online, round_trip, 
        Pigeons.timing_extrema, 
        Pigeons.allocation_extrema])
\end{lstlisting}
Note that to convert the Turing model into an appropriate Pigeons.jl target for sampling, 
we pass the model as an argument to \texttt{TuringLogPotential()}.
Once we have stored the PT output in the variable \texttt{pt} we can 
access the results, as described in the following section. 
The standard output after running the above code chunk is displayed in \cref{fig:standard_out}
and explained in the next section. 
For purposes of comparison, we also run a traditional 
(single-chain Markov chain Monte Carlo) method.
